---
title: "RepReschProject2 using the NOAA Storm Database to answer some questions about severe weather events"
author: "Christopher Stonell"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read data from .bz2 file
Data read from the provided .bz2 file. read.csv supports directly reading from compressed .bz2 files.  
data is read into the variable "data".  
```{r read the data, echo = TRUE, cache = TRUE}
# read.csv allows for reading directly from compressed .bz2
data <- as.data.frame(read.csv("/home/christopher/Documents/Coursera_Data_Science/datasciencecoursera/reproducible-research/RepData_PeerAssessment2/repdata_data_StormData.csv.bz2"))

```
The dataset consists of `r nrow(data)` rows and `r ncol(data)` variables.

## Pre-processing & data transformations
```{r}
# set colClasses for the columns of interest
data$EVTYPE <- as.character(data$EVTYPE)
data$FATALITIES <- as.numeric(data$FATALITIES)
data$INJURIES <- as.numeric(data$INJURIES)

```


## Data Processing
```{r}
# do computations for results in this block
# measures of harm to health are columns FATALITIES & INJUTRIES.
topFatalities <- as.data.frame(sort(tapply(data$FATALITIES, data$EVTYPE, FUN=sum), decreasing = TRUE))
topInjuries <- as.data.frame(sort(tapply(data$INJURIES, data$EVTYPE, FUN=sum), decreasing = TRUE))
topCombined <- as.data.frame(sort(tapply((data$FATALITIES+data$INJURIES), data$EVTYPE, FUN=sum), decreasing = TRUE))
d_fatal <- data.frame(x = topFatalities, group=names(topFatalities))
d_inj <- data.frame(x = topInjuries, group=names(topInjuries))
d_comb <- data.frame(x = topCombined, group=names(topCombined))
```

## results
### Questions 1. Across the United States, which types of events (as indicated in the EVTYPEEVTYPE variable) are most harmful with respect to population health?
The top 10 events with the greatest number fatalities were:  
`r knitr::kable(d_fatal[1:10,])`  

The top 10 events with the greatest number of injuries were:  
`r knitr::kable(d_inj[1:10,])`  

The harm to public heath can be considered measured by the sum of fatalities and injuries. The top 10 events with the greatest number of fatalities and injuries were:  
`r knitr::kable(d_comb[1:10,])`  


### Question 2. Across the United States, which types of events have the greatest economic consequences?

