---
title: RepReschProject2 using the NOAA Storm Database to answer some questions about
  severe weather events
author: "Christopher Stonell"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
In this analysis project I am using the data from the National Oceanic and Atmospheric Administration's Storm Database from 1950 to 2011.
I will address two questions in particular:  
1. What weather events have the greatest impact on public health
2. What weather events have the greatest impact in economic indicators.  
The analysis will be presented as tables and figures.

## Read data from .bz2 file
Data read from the provided .bz2 file. read.csv supports directly reading from compressed .bz2 files.  
The data is read as a data frame into the variable "data".  
```{r read the data, echo = TRUE, cache = TRUE}
# read.csv allows for reading directly from compressed .bz2
data <- as.data.frame(read.csv("/home/christopher/Documents/Coursera_Data_Science/datasciencecoursera/reproducible-research/RepData_PeerAssessment2/repdata_data_StormData.csv.bz2"))

```
The dataset consists of `r nrow(data)` rows and `r ncol(data)` variables.

## Pre-processing & data transformations
```{r}
# set colClasses for the columns of interest to ensure computations correct.
data$EVTYPE <- as.character(data$EVTYPE)
data$FATALITIES <- as.numeric(data$FATALITIES)
data$INJURIES <- as.numeric(data$INJURIES)
data$PROPDMG <- as.numeric(data$PROPDMG)
data$CROPDMG <- as.numeric(data$CROPDMG)

```


## Data Processing
```{r computations for results sections}
# do computations for results in this block

# measures of harm to health are columns FATALITIES & INJUTRIES.
topFatalities <- sort(tapply(data$FATALITIES, data$EVTYPE, FUN=sum), decreasing = TRUE)
topInjuries <- sort(tapply(data$INJURIES, data$EVTYPE, FUN=sum), decreasing = TRUE)
topCombined <- sort(tapply((data$FATALITIES+data$INJURIES), list(data$EVTYPE), FUN=sum), decreasing = TRUE)
d_comb <- data.frame(x = topCombined[1:10], group=names(topCombined[1:10]))

# measures of economic impact
topEconomic <- sort(tapply((data$PROPDMG+data$CROPDMG), list(data$EVTYPE), FUN=sum), decreasing = TRUE)

d_econ <- data.frame(x = topEconomic[1:10], group=names(topEconomic[1:10]))
```

## results
### Questions 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
The events are captured in the variable called EVTYPE. There are two variables for health impacts: FATALITIES and INJURIES. The harm to public heath can be considered measured by the sum of fatalities and injuries.  
The top 10 events with the greatest number of fatalities and injuries were:  
`r knitr::kable(d_comb, col.names=c("Number",""), caption = "Table of number of Fatalities and Injuries by Severe Weather Event type.")`  

### Question 2. Across the United States, which types of events have the greatest economic consequences?
The economic consequences are captured as dollar amounts in variables PROPDMG for cost of property damage and CROPDMG for cost of copr damage.  
The events with the top 10 greatest economic impacts were:
`r knitr::kable(d_econ, col.names=c("$",""), caption = "Table of economic cost of Severe Weather Events by Types.")`

### Bar Graph of economic impact by year
```{r barplot, echo=TRUE}
## use tapply again to generate the data for barplot by state
barplot(tapply((data$PROPDMG+data$CROPDMG), list(data$STATE), FUN=sum), xlab="State", ylab = "Total $ economic damage", main="Barplot of Total economic impact of severe weather events by State")
```